 

### 2021-07-28 vertical-thermal-review.Rmd
Edits in introduction section

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}$C by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilitiesa 2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilitiesa 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilitiesa; @mcdowellPervasiveShiftsForest2020]. Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006a].

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating forest macroclimates [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], as well as microclimates by buffering understories from extreme temperatures and other macroclimatic conditions [@zellweger_seasonal_2019].
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here, we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence $T_{leaf}$ and metabolism;
(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) size-structuring of whole-tree and ecosystem function.
*We then consider the implications for understanding forest responses to global change, including how these scale across space and time.*

@@

**The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs. `r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest itself.** 
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S. National Ecological Observatory Network [NEON; Figs. `r fig_NEON_vertical`-`r SIfig_NEON_all`; Appendix `r appendix_NEON_methods`, Schimel et al. 2007].
While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height (Parker 1995).

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six sites in the National Ecological Observatory Network (NEON)**. Shown are  height profiles in growing season (a) leaf area density, (b) proportion sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV). Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig. S1.](NEON_height_profiles/figures/profile_all.png)

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations followed by disturbance all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs. `r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].** 
This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than under more open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs. `r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`). 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

**Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999]. Elevated CO~2~ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.

Relative humidity, RH, tends to be higher in the understory and decreases with height across the vertical profile, although this trend is absent in open forests [Figs. `r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018].
Dense-canopy forests maintain higher daily maximum RH below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].Especially when canopy height and leaf area index are large, maximum Tair and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].

**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [@defrenneForestMicroclimatesClimate2021]. 
Typically, dense canopies buffer understories from high maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs. `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

@@

**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism (see following section) across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).

>>>>

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}$C by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including understory microclimates by buffering extreme temperatures, solar insolation, wind, and other meteorological conditions [@zellweger_seasonal_2019]. This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
This vertical gradient shapes nearly every aspect of plant metabolism, demography, and ecology; however, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here, we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence $T_{leaf}$ and metabolism;
(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) the size-structuring of whole-tree and ecosystem function.
*We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.*

@@

**The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs. `r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest.** 
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S. National Ecological Observatory Network [NEON; Figs. `r fig_NEON_vertical`-`r SIfig_NEON_all`; Appendix `r appendix_NEON_methods`, Schimel et al. 2007].
While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height (Parker 1995).

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six sites in the National Ecological Observatory Network (NEON)**. Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV). Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig. S1.](NEON_height_profiles/figures/profile_all.png)

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is structured in correspondence with the distribution of tree sizes and tree architecture within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns and further influence the formation of closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs. `r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].** 
This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than under open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs. `r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

**Wind speeds are also higher at the top of the canopy, owing to the effect of turbulence buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`). 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the vertical profile than the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for both water and heat loss (see next section).

**Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999]. Elevated CO~2~ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.

Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Figs. `r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018].
Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Especially when canopy height and leaf area index (LAI) are large, maximum Tair and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].

**Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
The $T_{air}$ gradient from the canopy to the understory varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation type [@defrenneForestMicroclimatesClimate2021]. 
Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs. `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

@@

**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
